---
title: "NYPD Shooting Incident Data (Historic)"
author: "Stuart Akeson"
date: "8/25/2021"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

# Neccesary Libraries:

```{r libraries, echo=TRUE, results='hide', warning=FALSE, message=FALSE}
library(tidyverse, warn.conflicts = F, quietly = T)
library(plyr)
library(gridExtra)
```

# Importing Data:
The NOAA data for average maximum temperature was accessed by using the NOAA data
search engine with the timeframe set for 1/1/2006 - 12/31/2020, and the Laguardia Airport
terminal was chosen as the representative indicator of weather. This is because 
the Laguardia station is centrally located between all of the boroughs which
will be featured later in this analysis.
```{r importing_data}
shooting_data_url = 
"https://data.cityofnewyork.us/api/views/
833y-fsy8/rows.csv?accessType=DOWNLOAD"
df_shooting = read.csv(shooting_data_url)
head(df_shooting, 3)

weather_data = "./Daily_weather_NYC.csv"
df_weather = read.csv(weather_data)
head(df_weather, 3)
```

# Setting Analysis Parameters:
This analysis will focus on the relationship between maximum temperature data
and the relationship it has with shooting incidents in New York City. We will
begin by removing columns that we're not going to be focusing on for this
analysis. We are only interested in the maximum temperature for each day, meaning
we can remove all the columns besides the date and the "TMAX" column. Similarly
for the shooting incident data we can view each row as a single incident, and beyond
that we are only concerned with the date it occured on so we can match it to the
relevant maximum temperature and the borough in which the shooting occured. Again
the borough data will be featured later in this analysis.

```{r cleaning_data_1}
df_weather = subset(df_weather, select = c(DATE, TMAX))
df_shooting = subset(df_shooting, select =c(OCCUR_DATE, BORO))
head(df_weather, 3)
head(df_shooting, 3)
```

# Updating data type of Date Columns:
We need to update the data type of the Date Columns to reflect that they are
indeed dates. This way we will be able to join the dataframes and continue
with our analysis. Here we are using a full join to create the single data frame.
A full join is chosen so that we can gather data about the total number of days of
each temperature in addition to the number of shootings that occurred at each
temperature.

```{r updating_data_types}
df_shooting$OCCUR_DATE = as.Date(df_shooting$OCCUR_DATE, format="%m/%d/%Y")
colnames(df_shooting)[1]<-"DATE"

df_weather$DATE = as.Date(df_weather$DATE, format="%Y-%m-%d")

merged_df = merge(x=df_shooting, y=df_weather, by="DATE", all=TRUE)
head(merged_df, 3)
```

# Calculating the number of shooting incidents per degree Farenheit:
We are going to calculate the total number of incidents that occurred per degree
Fahrenheit. Then we are going to have to normalize that number based on the number
of occurrences of the temperature in question.

```{r graphing preliminary findings}
shootings_by_temp <- count(merged_df, vars="TMAX")
colnames(shootings_by_temp)[2]<-"Shootings"

temp_counts <- count(df_weather, vars = "TMAX")
non_normalized_scatter_plot <- ggplot(shootings_by_temp, aes(x=TMAX, y=Shootings)) +
geom_point(color="blue") +
labs(title="Raw Shooting Incidents per Degree Farenheit")+
xlab("Maximum Temperature")+
ylab("Non Normalized Number of Shooting Incidents")
non_normalized_scatter_plot

temp_shooting_relative_df <- merge(x=shootings_by_temp, y=temp_counts, vars="TMAX", all=TRUE)
temp_shooting_relative_df <- mutate(.data=temp_shooting_relative_df, relative_freq=Shootings/freq)

relative_freq_scatter = ggplot(data= temp_shooting_relative_df, aes(x=TMAX, y=relative_freq))+
        geom_point(color="red")
relative_freq_scatter

```


# Modeling the relationship:
For this analysis we are going to assume that the relationship is linear, and we
are going to model relationship using the linear modeling least squares estimator
method. In R this is accomplished by the lm() function. The result will be a
linear regression line that best approximates the observed data.

```{r modeling relationship}
relative_frequency_model <- lm(relative_freq ~ TMAX, 
                               data= temp_shooting_relative_df)
summary(relative_frequency_model)
relative_freq_scatter + geom_smooth(method="lm")
```
We can see that there appears to be a linear trend in the data, but the upper tail
and the outlier on the lower end make a further investigation worth while.
We are going to use the basic R Plot function to asses the linearity of the data.
We will see that the data appears to be near linear, or linear in most cases,
but that the standardized residuals do not have a linear relationship with the
fitted data. If nothing else this should give us pause in drawing meaningful 
conclusions from any linear model applied to this data set.
```{r Model Linearity}
par(mfrow= c(2,2))
plot(relative_frequency_model)
```

# Analysis by Borough and Maximum Temperature:
The variability in the numbers of shootings that occur in each borough lead to the
natural question of whether that are more statistically significant trends in each
of the seperate boroughs, or if the trend falls apart when the data is split into 
the component boroughs. The analysis will proceed in much the same way, except this
time there will be 5 different models produced, one for each borough.
```{r borough temperature analysis, warning=FALSE, message=FALSE}
shootings_by_temp_and_borough <- count(merged_df, vars=c("TMAX", "BORO"))
colnames(shootings_by_temp_and_borough)[3]<-"Shootings"
shootings_by_temp_and_borough_with_count <- merge(
        x=shootings_by_temp_and_borough, y=temp_counts, vars="TMAX", all=TRUE)
shootings_by_temp_and_borough_relative_frequency <- mutate(
        .data=shootings_by_temp_and_borough_with_count, 
        relative_freq=Shootings/freq)

#Here we are going to calculate all 5 of the linear models:
brooklyn_model = lm(relative_freq ~ TMAX, 
        data=shootings_by_temp_and_borough_relative_frequency[
        shootings_by_temp_and_borough_relative_frequency$BORO=="BROOKLYN",])

manhattan_model = lm(relative_freq ~ TMAX, 
        data=shootings_by_temp_and_borough_relative_frequency[
        shootings_by_temp_and_borough_relative_frequency$BORO=="MANHATTAN",])

bronx_model = lm(relative_freq ~ TMAX, 
        data=shootings_by_temp_and_borough_relative_frequency[
        shootings_by_temp_and_borough_relative_frequency$BORO=="BRONX",])

queens_model = lm(relative_freq ~ TMAX, 
        data=shootings_by_temp_and_borough_relative_frequency[
        shootings_by_temp_and_borough_relative_frequency$BORO=="QUEENS",])
staten_island_model = lm(relative_freq ~ TMAX, 
        data=shootings_by_temp_and_borough_relative_frequency[
        shootings_by_temp_and_borough_relative_frequency$BORO=="STATEN ISLAND",])
        
#And represent them visually:
brooklyn_graph =
ggplot(data=shootings_by_temp_and_borough_relative_frequency[
        shootings_by_temp_and_borough_relative_frequency$BORO=="BROOKLYN",], 
        aes(x=TMAX, y=relative_freq)) +
        geom_point(color="gold", shape=18) +
        geom_smooth(method="lm") +
        labs(title="Brooklyn") +
        xlab("Maximum Temperature") +
        ylab("Relative Freq.")+
        ylim(0, 5)

manhattan_graph =
ggplot(data=shootings_by_temp_and_borough_relative_frequency[
        shootings_by_temp_and_borough_relative_frequency$BORO=="MANHATTAN",], 
        aes(x=TMAX, y=relative_freq)) +
        geom_point(color="black", shape=18) +
        geom_smooth(method="lm") +
        labs(title="Manhattan") +
        xlab("Maximum Temperature") +
        ylab("Relative Freq.")+
        ylim(0, 5)
        
bronx_graph =
ggplot(data=shootings_by_temp_and_borough_relative_frequency[
        shootings_by_temp_and_borough_relative_frequency$BORO=="BRONX",], 
        aes(x=TMAX, y=relative_freq)) +
        geom_point(color="red", shape=18) +
        geom_smooth(method="lm") +
        labs(title="Bronx") +
        xlab("Maximum Temperature") +
        ylab("Relative Freq.")+
        ylim(0, 5)

queens_graph =
ggplot(data=shootings_by_temp_and_borough_relative_frequency[
        shootings_by_temp_and_borough_relative_frequency$BORO=="QUEENS",], 
        aes(x=TMAX, y=relative_freq)) +
        geom_point(color="green", shape=18) +
        geom_smooth(method="lm") +
        labs(title="Queens") +
        xlab("Maximum Temperature") +
        ylab("Relative Freq.")+
        ylim(0, 5)

staten_island_graph=
ggplot(data=shootings_by_temp_and_borough_relative_frequency[
        shootings_by_temp_and_borough_relative_frequency$BORO=="BROOKLYN",], 
        aes(x=TMAX, y=relative_freq)) +
        geom_point(color="blue", shape=18) +
        geom_smooth(method="lm") +
        labs(title="Staten Island") +
        xlab("Maximum Temperature") +
        ylab("Relative Freq.") +
        ylim(0, 5)

grid.arrange(brooklyn_graph, bronx_graph, manhattan_graph, 
             queens_graph, staten_island_graph, ncol=3, nrow=2)
```

#Bias and Conclusions:

One of the first pieces of bias to mention is in the representation of the data visually in the
final grid of 5 graphs. There is a data point for the Bronx Borough at a temperature of 15 degrees
Fahrenheit with 19 shooting incidents. This falls well outside of the chosen y axis dimensions. It still
has an impact on the regression line, but it does not appear on the graph. 
The next source of bias to consider is the viability of splitting the data into the relative Boroughs.
While the regression lines for some of the boroughs may be a tighter fit to the data than the general model,
it means that none of these regressions can be used to predict or extrapolate for any of the other boroughs.
This being said it is worth noting that there is a positive linear correlation for all 5 boroughs when 
looking at the relative frequency of shooting incidents and the the maximum temperature recorded for those 
dates at Laguardia airport. 
Another source of bias is that Laguardia airport might not accurately reflect the temperature in each
borough, it was chosen for its central location, but that does not make it the perfect measurement for 
the true temperature in each borough.

```{r R_squared}
summary(queens_model)
summary(bronx_model)
summary(brooklyn_model)
summary(manhattan_model)
summary(staten_island_model)
```

```{r session info}
sessionInfo()
```